---
title: "Baltimore Environmental Citations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
library(readxl)
library(lubridate)
library(janitor)
library(tidyverse)
library(ggplot2)
library(sf)
```

## Load in data

```{r}
citations <- read_csv('data/Environmental_Citations.csv')
```

## Clean names

```{r}
citations <- citations %>% 
  clean_names()
```

## Clean, go from citation neighborhoods -> CSAs

```{r}
citations_geo <- citations %>%
  separate(violation_date, c("violation_date", "violation_time"), " ") %>%
  mutate(violation_date = ymd(violation_date),
         violation_year = year(violation_date),
         violation_month = month(violation_date),
         violation_day = day(violation_date),
         location = gsub("\\(", "", location),
         location = gsub("\\)", "", location)) %>%
  separate(location, c("lat", "lon"), ",") %>% 
  filter(!is.na(block)) %>%
  filter(!is.na(lon)) %>%
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326)

citations_invalid <- citations %>%
  mutate(location = gsub("\\(", "", location),
         location = gsub("\\)", "", location)) %>% 
  separate(location, c("lat", "lon"), ",") %>% 
  filter(is.na(block) | is.na(lon))

csas <- read_sf("data/Community_Statistical_Areas_(2020)/Community_Statistical_Areas__2020_.shp") %>%
  st_transform(crs = 4326) %>%
  clean_names()

# citations_by_csas <- st_intersection(geocoded_citations,csas)
# write_sf(citations_by_csas, "data/geocoded-citations-by-csa.csv")

env_cit <- read_sf("data/geocoded-citations-by-csa.csv") %>%
  clean_names()
```

## Filter to 2015 onwards

```{r}
env_cit_new <- env_cit %>% 
  filter(violation_year >= 2012)
```

## Read in Demographics

```{r}
demos <- read_csv('data/2020-demos-bmore.csv') %>%
  clean_names() %>% 
  select(-c(total_male_population,total_female_population)) %>% 
  mutate(across(is.numeric,~round(., digits=1))) %>% 
  rename(csa2020 = csa2010)
```

## Join datasets

```{r}
env_cit_demos <- left_join(env_cit_new, demos, by='csa2020')
```

## Citations by neighborhood with demos

```{r}
cit_by_race_csa <- env_cit_demos %>% 
  group_by(csa2020,total_population,percent_of_residents_black_african_american_non_hispanic,percent_of_residents_white_caucasian_non_hispanic,percent_of_residents_asian_non_hispanic,percent_of_residents_two_or_more_races_non_hispanic,percent_of_residents_hispanic,median_household_income) %>% 
  summarize(count_cit=n()) %>% 
  mutate(cit_per_capita = round((count_cit/total_population),digits=2)) %>% 
  arrange(desc(cit_per_capita))

csa_export <- cit_by_race_csa %>% 
  rename(
    black = percent_of_residents_black_african_american_non_hispanic,
    white = percent_of_residents_white_caucasian_non_hispanic,
    asian = percent_of_residents_asian_non_hispanic,
    mixed = percent_of_residents_two_or_more_races_non_hispanic,
    hispanic = percent_of_residents_hispanic
  ) %>% 
  write_csv("data/csa_cit_data.csv")
```

# Neighborhood Analysis

```{r}
start_date <- as_date('2012-01-01')
end_date <- as_date('2023-01-01')

int <- interval(start_date,end_date)

month_seq = seq(start_date, end_date, by='month')
```

## Clifton-Berea

```{r}
clifton_berea <- env_cit_demos %>% 
  filter(neighborhood == 'Clifton-Berea')

clifton_berea_by_desc <- clifton_berea %>% 
  group_by(description) %>% 
  summarise(count=n())

clifton_berea_over_time <- clifton_berea %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(clifton_count=n())

ggplot(data = clifton_berea_over_time) +
  geom_area(mapping = aes(x = month, y = clifton_count))
```

## Midway/Coldstream

```{r}
midway_coldstream <- env_cit_demos %>% 
  filter(neighborhood == 'Midway/Coldstream')

midway_coldstream_by_desc <- midway_coldstream %>% 
  group_by(description) %>% 
  summarise(count=n())

midway_coldstream_over_time <- midway_coldstream %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(midway_count=n())

ggplot(data = midway_coldstream_over_time) +
  geom_area(mapping = aes(x = month, y = midway_count))
```

## Belair-Edison

```{r}
belair_edison <- env_cit_demos %>% 
  filter(neighborhood == 'Belair-Edison')

belair_edison_by_desc <- belair_edison %>% 
  group_by(description) %>% 
  summarise(count=n())

belair_edison_over_time <- belair_edison %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(belair_count=n())

ggplot(data = belair_edison_over_time) +
  geom_area(mapping = aes(x = month, y = belair_count))
```

## Madison/East End

```{r}
madison_east_end <- env_cit_demos %>% 
  filter(neighborhood == 'Madison/East End')

madison_east_end_by_desc <- madison_east_end %>% 
  group_by(description) %>% 
  summarise(count=n())

madison_east_end_over_time <- madison_east_end %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(madison_count=n())

ggplot(data = madison_east_end_over_time) +
  geom_area(mapping = aes(x = month, y = madison_count))
```

## Southwest Baltimore

```{r}
sw_bmore <- env_cit_demos %>% 
  filter(neighborhood == 'Southwest Baltimore')

sw_bmore_by_desc <- sw_bmore %>% 
  group_by(description) %>% 
  summarise(count=n())

sw_bmore_over_time <- sw_bmore %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(sw_count=n())
ggplot(data = sw_bmore_over_time) +
  geom_area(mapping = aes(x = month, y = sw_count))
```

## South Baltimore

```{r}
s_bmore <- env_cit_demos %>% 
  filter(neighborhood == 'South Baltimore')

s_bmore_by_desc <- s_bmore %>% 
  group_by(description) %>% 
  summarise(count=n())

s_bmore_over_time <- s_bmore %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(s_count=n())

ggplot(data = s_bmore_over_time) +
  geom_area(mapping = aes(x = month, y = s_count))
```

## Canton

```{r}
canton <- env_cit_demos %>% 
  filter(neighborhood == 'Canton')

canton_by_desc <- canton %>% 
  group_by(description) %>% 
  summarise(count=n())

canton_over_time <- canton %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(canton_count=n())

ggplot(data = canton_over_time) +
  geom_area(mapping = aes(x = month, y = canton_count))
```

## Greater Roland Park/Poplar Hill

```{r}
roland_park <- env_cit_demos %>% 
  filter(neighborhood == 'Greater Roland Park/Poplar Hill')

roland_park_by_desc <- roland_park %>% 
  group_by(description) %>% 
  summarise(count=n())

roland_park_over_time <- roland_park %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(roland_park_count=n())

ggplot(data = roland_park_over_time) +
  geom_area(mapping = aes(x = month, y = roland_park_count))
```

## Inner Harbor/Federal Hill

```{r}
inner_harbor <- env_cit_demos %>% 
  filter(neighborhood == 'Inner Harbor/Federal Hill')

inner_harbor_by_desc <- inner_harbor %>% 
  group_by(description) %>% 
  summarise(count=n())

inner_harbor_over_time <- inner_harbor %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(inner_harbor_count=n())

ggplot(data = inner_harbor_over_time) +
  geom_area(mapping = aes(x = month, y = inner_harbor_count))
```

## Medfield/Hampden/Woodberry/Remington

```{r}
mhwr <- env_cit_demos %>% 
  filter(neighborhood == 'Medfield/Hampden/Woodberry/Remington')

mhwr_by_desc <- mhwr %>% 
  group_by(description) %>% 
  summarise(count=n())

mhwr_over_time <- mhwr %>% 
  group_by(month=floor_date(violation_date,'month')) %>% 
  summarise(mhwr_count=n())

ggplot(data = mhwr_over_time) +
  geom_area(mapping = aes(x = month, y = mhwr_count))
```

## Comparative graph - most cited vs. whitest

```{r}
month_seq = seq(start_date, end_date, by='month')

ten_over_time <- data.frame('month' = month_seq) %>% 
  left_join(clifton_berea_over_time,by = 'month') %>% 
  left_join(midway_coldstream_over_time,by = 'month') %>% 
  left_join(belair_edison_over_time,by = 'month') %>% 
  left_join(madison_east_end_over_time,by = 'month') %>% 
  left_join(sw_bmore_over_time,by = 'month') %>% 
  left_join(s_bmore_over_time,by = 'month') %>% 
  left_join(canton_over_time,by = 'month') %>% 
  left_join(roland_park_over_time,by = 'month') %>% 
  left_join(inner_harbor_over_time,by = 'month') %>% 
  left_join(mhwr_over_time,by = 'month')

ten_over_time[is.na(ten_over_time)] = 0

ggplot(data = ten_over_time, aes(x = month)) +
  geom_line(aes(y=clifton_count,color='Majority Black')) +
  geom_line(aes(y=midway_count,color='Majority Black')) +
  geom_line(aes(y=belair_count,color='Majority Black')) +
  geom_line(aes(y=madison_count,color='Majority Black')) +
  geom_line(aes(y=sw_count,color='Majority Black')) +
  geom_line(aes(y=s_count,color='Majority white')) +
  geom_line(aes(y=canton_count,color='Majority white')) +
  geom_line(aes(y=roland_park_count,color='Majority white')) +
  geom_line(aes(y=inner_harbor_count,color='Majority white')) +
  geom_line(aes(y=mhwr_count,color='Majority white')) +
  labs(x = 'month', y = 'citations issued', title = str_wrap('Majority Black neighborhoods received far more environmental citations over the past ten years than majority white neighborhoods.', 70), color = 'Neighborhood type')
  
```

## Scatterplot - Median income vs. citations per capita

```{r}
ggplot(data = cit_by_race_csa) +
  geom_point(aes(x = median_household_income, y = cit_per_capita, color = percent_of_residents_black_african_american_non_hispanic, size = count_cit)) +
  geom_smooth(method = lm,aes(x = median_household_income, y = cit_per_capita)) +
  labs(x = 'Median household income', y = 'Environmental citations issued per capita', title = str_wrap('Poorer neighborhoods receive more citations per capita than richer neighborhoods.', 70), color = 'Percent of residents who are Black', size = 'Total citations issued since 2012')
```

## Most cited addresses

```{r}
cit_by_add <- env_cit_new %>% 
  group_by(violation_location) %>% 
  summarise(count_cit=n()) %>% 
  arrange(desc(count_cit))
```


## High cit count addresses

```{r}
top_adds <- cit_by_add %>% 
  filter(
    str_detect(violation_location,"^[0-9][0-9][0-9][0-9]") &
    !str_detect(violation_location, "^0000")
  ) %>% 
  head(5)

top_adds
```

#Vacant housing

```{r}
vacant_housing <- env_cit_demos %>% 
  filter(description == "FAILURE TO FILE ANNUAL VACANT BUILDING REGISTRATION") %>%
  group_by(neighborhood) %>% 
  summarise(count = n())
```

# Export csa
```{r}
cit_by_race_csa <- cit_by_race_csa %>% 
  mutate(majority_black = case_when(
    percent_of_residents_black_african_american_non_hispanic > 50 ~ "Yes",
    TRUE ~ "No"
  ))

write_csv(cit_by_race_csa, path = "data/csa_cit_data_fixed.csv")
```

# Total cost of cits by neighborhood

```{r}
cost_by_hood <- env_cit_demos %>% 
  mutate(majority_black = case_when(
    percent_of_residents_black_african_american_non_hispanic > 50 ~ "Yes",
    TRUE ~ "No"
  )) %>% 
  filter(violation_year == 2022) %>% 
  group_by(csa2020,majority_black,median_household_income) %>%
  mutate(fine_amount = as.numeric(fine_amount)) %>% 
  summarise(
    total_cost = sum(fine_amount)
  ) %>% 
  arrange(desc(total_cost))

cost_by_hood <- cost_by_hood[(1:15),]

write_csv(cost_by_hood, file="data/cost_by_hood.csv")
```

## Vacant housing

```{r}
housing <- env_cit_new %>%
  filter(description == "FAILURE TO FILE ANNUAL VACANT BUILDING REGISTRATION") %>% 
  group_by(csa2020) %>% 
  summarise(count = n()) %>% 
  arrange(count)
```

